---
title: "Tutorial"
output: html_document
bibliography: references.bib
link-citations: TRUE
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, include = FALSE}
library(ecorisk)
```

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

This tutorial walks you through a complete risk assessment using the
ecorisk package with provided demo datasets. For background on the
underlying framework and theory, see the [Background & Framework](ecorisk-background.html) article.

# Scenario

Let's suppose we want to conduct a risk assessment for a fictional
marine ecosystem. We want to investigate whether the system is at risk
of being in an undesired state due to 5 ongoing pressures:

-   Temperature increase due to climate change
-   Salinity decrease due to climate change
-   A decreasing nutrient input after a phase of eutrophication
-   Oxygen depletion
-   Fishing pressure

To best represent our ecosystem we choose state indicators from
different trophic levels: phytoplankton, zooplankton, two fish species
(herring and cod), and seabirds. For some of these we have more or less
detailed expert knowledge and for some of them we have time series
representing this state indicator. Thus we decide on the following
assessment scheme using both the expert scoring and the modelling
pathway of ecorisk:

```{r indicator overview, echo = FALSE}
ind_data <- data.frame(
  Trait = c("Cod", " Herring"), 
  General = c("Phytoplankton", "Seabirds"),
  Model = c("Cod spawning stock biomass", "Zooplankton mean size")
)

knitr::kable(ind_data, booktabs = TRUE) |>
  kableExtra::add_header_above(c("Expert scoring pathway" = 2, 
    "Modelling pathway" = 1)) 
```

For the two fish species enough knowledge is available for a detailed
expert scoring using species traits. Phytoplankton and seabirds are
larger species groups, a scoring on a trait basis would be very
difficult, therefore the experts will give here only one general score
for sensitivity and adaptive capacity of these two groups. The time
series we want to use for the modeling approach are i) spawning stock
biomass of the cod stock as indicator for the health of the species and
ii) zooplankton mean size which represents the status of the zooplankton
group and gives an indication about the status of the food web. The
pressures will be assessed using both pathways.

# Data preparation

First step is to create the scoring tables, which will be filled out by
our experts. With the functions `create_template_exposure()` and
`crt_sensitivity()`, we can automatically create tables for further
usage in the ecorisk workflow. The functions need the names of the
pressures and for sensitivity also the names of the indicators. For
exposure we want to investigate 4 components:

-   The magnitude of change,

-   the future trend,

-   the frequency of the change and

-   the spatial coverage of the change in the ecosystem.

To define vulnerability the experts should score the sensitivity and the
adaptive capacity. They should differentiate between two types of
effect: direct effects only and the combination of direct and indirect
effects. Indirect effects occur for example due to food web
interactions. If possible the experts will score four individual species
traits:

-   Feeding

-   Behaviour

-   Reproduction

-   Habitat

otherwise they will give general sensitivity and adaptive capacity
scores. Both the exposure and the sensitivity scoring include an
uncertainty assessment. As a second step we rename the variables in our
scoring templates.

```{r create tables, class.output="scroll-100"}
exposure_scoring <- create_template_exposure(
  pressures = c("temperature", "salinity", "oxygen", "nutrient", "fishing"),
  n_components = 4, 
  mode_uncertainty = "component"
)
names(exposure_scoring)
# Rename exposure components
names(exposure_scoring)[2:9] <- c("magnitude", "frequency", "trend", "spatial",
                                  "uncertainty_magnitude", "uncertainty_frequency",
                                  "uncertainty_trend", "uncertainty_spatial")

sensitivity_scoring <- create_template_sensitivity(
  indicators = c("phytoplankton", "herring", "cod", "seabirds"),
  pressures = c("temperature", "salinity", "oxygen", "nutrient", "fishing"),
  type = c("direct", "direct_indirect"),
  n_sensitivity_traits = 5,
  mode_adaptive_capacity = "trait",
  mode_uncertainty = "trait"
)
names(sensitivity_scoring)

# Replace the generic trait names ('...trait_1', '...trait_2') with
# the actual trait names
names(sensitivity_scoring) <- names(sensitivity_scoring) |> 
  stringr::str_replace("trait_1$", "feeding") |> 
  stringr::str_replace("trait_2$", "behaviour") |> 
  stringr::str_replace("trait_3$", "reproduction") |> 
  stringr::str_replace("trait_4$", "habitat") |> 
  stringr::str_replace("trait_5$", "general") 
```

The experts have filled out the scoring tables in a joint exercise, thus
we do not have to aggregate the individual expert scores. If individual
expert scores are given, it is recommended to follow the work flow until
the risk calculation and aggregation of risk scores is completed and
then aggregate these scores across all experts. This allows for more
precise and detailed results, as differences in the results between
experts can be evaluated.

```{r read expert scores}
exposure_scoring <- ex_expert_exposure
head(exposure_scoring)
sensitivity_scoring <- ex_expert_sensitivity
head(sensitivity_scoring)
```

The experts followed this scoring scheme:

-   Exposure 1–5 (low to high)

-   Sensitivity −5–5 (high negative to high positive influence)

-   Adaptive capacity −1–1 (low to high)

-   Uncertainty 1–3 (low to high)

For more information about this scheme see also the [Scoring](ecorisk-background.html#scoring) section in the Background article.
This scoring scheme aligns with the results of the modelling pathway.

To prepare the modelling pathway we load our time series data. The data
includes indicator variables for our five pressures and two state
indicators (cod and zooplankton). The data covers a time frame from 1984
to 2016. The data was created based on trends of similar indicators from
the Baltic Sea. There is another example data set with data from the
North Sea.

```{r read time series data}
ts_pressures <- pressure_ts_baltic
head(ts_pressures)
ts_indicators <- indicator_ts_baltic
head(ts_indicators)
```

# Analysis

Now we will calculate exposure and sensitivity scores in both pathways.
We start with the expert scoring pathway using the functions
`calc_exposure()` and `calc_sensitivity()`. They calculate from the
exposure components and sensitivity trait scores general scores using an
aggregation method selected by the user. We will use the arithmetic mean
to aggregate component and trait scores. Other options are median,
minimum or maximum.

```{r exposure and sensitivity expert pathway, echo = TRUE}
exp_expert <- calc_exposure(
  pressures = exposure_scoring$pressure, 
  components = exposure_scoring[ ,2:5],
  uncertainty = exposure_scoring[ ,6:9],
  method = "mean"
)
head(exp_expert)

sens_ac_expert <- calc_sensitivity(
  indicators = sensitivity_scoring$indicator, 
  pressures = sensitivity_scoring$pressure,
  type = sensitivity_scoring$type,
  sensitivity_traits = sensitivity_scoring[ ,4:8],
  adaptive_capacities = sensitivity_scoring[ ,9:13],
  uncertainty_sens = sensitivity_scoring[ ,14:18],
  uncertainty_ac = sensitivity_scoring[ ,19:23], 
  method = "mean"
)
head(sens_ac_expert)
```

Both data sets now contain the aggregated expert scores. The sensitivity
dataset also contains the initial expert scores, thus we can still
calculate vulnerability for each trait individually, which will lead to
more precise results. Additionally the last column gives information
about the pathway of the scores to compare later on the results from
both pathways.

In the modelling pathway we use the functions `model_exposure()` and
`model_sensitivity()`. To assess exposure based on a time series we have
to define a period where baseline conditions are assumed, the baseline
conditions are then compared to the conditions in the assessment time
period. In this tutorial the baseline is set to the first 10 years of
the time series and the assessment time period are the last 5 years of
the time series. Since baseline conditions are considered as "good" for
our ecosystem all pressures should return to baseline conditions, we
specify this with the argument trend. The spatial scale cannot be
assessed with temporal data only, thus we have to specify the scores. We
use here the same value for the spatial component that the experts gave
for the pressures in the expert scoring.

```{r model exposure}
exposure_model <- model_exposure(
  pressure_time_series = ts_pressures,
  base_years = c(start = 1984, end = 1993),
  current_years = c(start = 2007, end = 2016),
  trend = "return", 
  spatial = c(2, 2, 5, 5, 3, 3, 2, 2)
)
exposure_model
```

The output gives us the aggregated exposure score, which is the
arithmetic mean of all exposure components. These are the same
components that have been evaluated by our experts. The
`model_exposure()` function does not evaluate the associated
uncertainty.

We should check for all pressures if the direction is capturing the
correct dynamics and not only a short term fluctuation.

To assess sensitivity with time series data we also have to specify the
assessment time period to determine if the pressure has in the
assessment time period a negative or positive influence on the state
indicator. The influence is determined using a simple linear model, so
we can simply check if the direction was correctly determined using a
simple plot. First we set up a data frame specifying for each indicator
pressure combination the assessment time period.

```{r model sensitivity}
sens_ac_model <- model_sensitivity(
  indicator_time_series = ts_indicators, 
  pressure_time_series = ts_pressures,
  current_years = c(start = 2010, end = 2016)
)
sens_ac_model
```

The output of the function contains for each indicator and pressure
combination the type of effect, the sensitivity score, the uncertainty
associated with the uncertainty scoring and the model parameters that
are used for the sensitivity scoring. The function evaluates sensitivity
using a generalized additive model, if it is significant the $R^2$ value
is used to set a score between 1 and 5. For non-significant models the
score is set to 0. The edf score evaluates non-linearity of the
relationships, as these are more risky, the sensitivity will be
increased for edf scores > 1, but maximum to 5.

The scoring of sensitivity heavily relies on generalized additive models,
therefore it is important to inspect model diagnostics of the applied
models. The function `plot_diagnostic_sensitivity()` creates for each
state and pressure indicator combination four diagnostic plots.

```{r model diagnostics sensitivity}
plot_diagnostic_sensitivity(
   indicator_time_series = ts_indicators[, c(1:2)],
   pressure_time_series = ts_pressures[, c(1:2)]
)
```

Let's continue by calculating the vulnerability for both pathways with
the function `vulnerability()`. The vulnerability function combines
exposure, sensitivity and adaptive capacity into the vulnerability
score. For negative sensitivity score the following function applies:

$V = (S + AC) - E$ and for positive scores:

$V = (S + AC) + E$,

where V = vulnerability, S = sensitivity, AC = adaptive capacity and E =
exposure. For trait based scoring this will be done for each trait
individually. The vulnerabilities of each trait will then be aggregated;
we can decide with the parameter `method_v` if we want to use an
arithmetic mean, median, minimum or maximum for the aggregation, same
applies for aggregation of trait based uncertainty scores.

```{r vulnerability}
vuln_experts <- vulnerability(
  exposure_results = exp_expert, 
  sensitivity_results = sens_ac_expert,
  method_vulnerability = "mean",
  method_uncertainty = "mean"
)

vuln_model <- vulnerability(
  exposure_results = exposure_model, 
  sensitivity_results = sens_ac_model,
  method_vulnerability = "mean",
  method_uncertainty = "mean"
)

head(vuln_experts)
head(vuln_model)
```

The output of the vulnerability gives us for each indicator pressure
type combination the vulnerability score with its associated uncertainty
and the pathway with which the vulnerability has been determined.

To calculate the risk we need to know whether the indicator is currently
in a good or undesired state, since the risk to be in an undesired state
is higher if the indicator is already in it. Status can be determined by
the experts or using the time series of the indicator. Our experts have
assessed that phytoplankton and seabirds are currently in a good state,
but herring and cod unfortunately not. An undesired state gets a score
of −1 and a good state of +1.

```{r status experts}
status_experts <- data.frame(
  indicator = c("phytoplankton", "herring", "cod", "seabirds"), 
  status = c("good", "undesired", "undesired", "good"),
  score = c(1, -1, -1, 1)
)
status_experts
```

In the modelling pathway the function `status()` evaluates indicator
status based on time series. To assess the status we have to set
baseline conditions where the indicator is assumed to be in a good
status and our assessment time frame for which the status shall be
evaluated (similar to the `model_exposure()` function). Additionally we
can set a range around the baseline mean which is considered as good
status. We can specify whether it is undesired if the indicator is in
the assessment time period below or above this range using the arguments
`sign` and `condition`. For our status assessment we assume good
baseline conditions in the first ten years as the mean ± 1 standard
deviation. Undesired status is defined as being below 1 standard
deviation so we set `sign = "-"` and `condition = "<"`.

```{r status model}
status_model <- status(
  indicator_time_series = ts_indicators,
  base_years = c(start = 1984, end = 1993),
  current_years = c(start = 2012, end = 2016),
  sign = "-", 
  condition = "<"
)
status_model
```

Both of our indicators are in the assessment time period in an undesired
status.

Now we combine vulnerability and status to the final risk scores. Both
pathways use the function `risk()`.

```{r risk}
risk_expert <- risk(
  vulnerability = vuln_experts, 
  status = status_experts
)
risk_model <- risk(
  vulnerability = vuln_model, 
  status = status_model
)

head(risk_expert)
head(risk_model)
```

The output of the risk function gives us per indicator–pressure–type
combination the risk, calculated from vulnerability and status, as well
as the uncertainty associated with the risk evaluation. Before we
combine both data frames we will rename the pressure variables of the
model based risk assessment to align with the names of the expert based
assessment. This is necessary to correctly aggregate the risk values per
pressure to multi pressure risk scores. Furthermore we will select for
each pressure, which has more than one variable, one of them. We follow
here a precautionary approach, thus we will select the one posing the
highest risk.

```{r rename and select pressure variables model pathway}
risk_model <- risk_model[c(1, 3, 5, 7, 8, 9, 12, 14:16), ]
risk_model$pressure <- c(
  "nutrient", "temperature", "salinity", "oxygen", "fishing",   # for zooplankton
  "nutrient", "temperature", "salinity", "oxygen", "fishing")   # for cod
```

Now we combine both risk assessment data sets and aggregate them to get
multi pressure, multi indicator and ecosystem risk scores. The
aggregated score will automatically be calculated per pathway, type,
pathway and type, and without accounting for type or pathway.

```{r aggregate pathways and risks}
risks <- rbind(risk_expert, risk_model)

aggregated_risk <- aggregate_risk(
  risk_results = risks, 
  method = "mean"
)

aggregated_risk$multi_indicator_risk
aggregated_risk$multi_pressure_risk
```

The aggregated risk scores are in three lists:

-   One with all risks aggregated per pressure,

-   one aggregated per indicator and

-   one with the ecosystem risk, which is an aggregation of all multi
    pressure risk scores.

In all lists the scores are first aggregated per type and pathway, then
per type regardless of the pathway, then per pathway regardless of the
type and as last regardless of the type and pathway.

# Results

To get an easier overview of the results and help with the
interpretation ecorisk provides two plotting functions `plot_radar()`
and `plot_heatmap()`.

The first one creates for each indicator a radar plot with the risk
scores per pressure and type. In the centre it shows the aggregated
multi-pressure risk; we have to select which we want to use. This plot
helps to identify per indicator the most influencing pressures and
compare results from different types. If uncertainty has been assessed
this will be displayed as a ring around the radar plot. In this example
we will use the aggregated score of direct and indirect effects,
regardless of the pathway. These are assumed to better reflect reality
compared to direct effects only.

```{r plot radar, fig.width=6, fig.height=6}
p_radar <- plot_radar(
  risk_scores = risks,
  aggregated_scores = aggregated_risk,
  type = "direct_indirect", 
  pathway = "combined"
)

p_radar[[1]]
p_radar[[2]]
p_radar[[3]]
p_radar[[4]]
p_radar[[5]]
p_radar[[6]]
```

All of the indicators have a negative multi-pressure score. The
indicators assessed with the expert scoring pathway show a higher risk
due to the combined direct and indirect effects compared to direct
effects only. The uncertainty is generally lower in the modelling
pathway. Comparing the modelling and expert pathway for cod it is
noticeable that both have a similar multi-pressure score but the
individual risk scores are quite different: The experts assessed fishing
as the most influencing pressure, whereas in the modelling pathway
salinity is considered as the pressure with the highest impact. To
directly compare modelling and expert pathway for cod we can correlate
them in a plot:

```{r correlation plot, fig.width=6, fig.height=6}
temp <- risks[c(26:30, 45:49), c(1, 2, 7)]
temp <- temp |>
  tidyr::pivot_wider(names_from = indicator, values_from = risk)

ggplot2::ggplot(dat = temp, 
    ggplot2::aes(x = cod, y = eastern_baltic_cod, colour = pressure)) +
  ggplot2::geom_point() +
  ggplot2::geom_abline(slope = 1, intercept = 0) +
  ggplot2::xlim(-10, 10) + 
  ggplot2::ylim(-10,10) +
  ggplot2::labs(x = "Expert-based pathway", y = "Modelling-based pathway") +
  ggplot2::theme_minimal()
```

We can see that the risk scores for nutrients, oxygen and temperature
are quite similar for both pathways. But the values for salinity
indicate no risk in the modelling pathway. This might be due to a
salinity time series that does not catch the ongoing dynamics, and thus
resulting in a non-significant relationship between pressure and
indicator. The scores for fishing have even opposing directions; this
could be due to a different understanding of fishing and the future
developments: In the modelling pathway it has been shown that fishing is
decreasing, which will be positive for the indicators, thus in the
future it will have a positive effect. But the expert probably assumed
that fishing will always have negative effects, without considering the
positive development of the decreasing fishing pressure. This is a
linguistic uncertainty, which should have been discussed in the planning
and scoping phase of the assessment, to ensure that all have a common
understanding. Nevertheless we should reiterate with our experts as well
as check if the `model_sensitivity()` and `model_exposure()` functions
have correctly determined the effect direction.

The function `plot_heatmap()` creates for each assessment type a heatmap
of all pressure indicator combinations, optionally with the associated
uncertainty. The aggregated multi-pressure and indicator scores are
shown at the bottom and to the left of the heatmap as well as the
system risk score at the bottom right. For the aggregated scores we have
to choose which one we want to use as we already did for the
`plot_radar()` function. This function gives a complete overview of the
system dynamics allowing to easily compare ecosystems and determine by
which indicators and pressures ecosystem risk is driven.

```{r plot heatmap, fig.width=8, fig.height=9}
p_heat <- plot_heatmap(
  risk_scores = risks,
  aggregated_scores = aggregated_risk,
  uncertainty = TRUE
)

p_heat[[1]]
p_heat[[2]]
```

The first heatmap includes only direct effects, thus this heatmap is a
bit smaller as we did not assess for all indicators direct effects only.
The multi pressure and indicator scores are the aggregated direct scores
for all assessed pathways. The grey frame around the heatmap tiles show
the uncertainty associated with the score. We can see with this heatmap
that cod is the indicator most at risk, this risk is due to high effects
of fishing and temperature. The least at risk indicators are
phytoplankton and seabirds. Both are not affected by several of the
pressures, but still seabirds are at a medium to high risk due to
fishing. The second heatmap shows the direct and indirect effects
including also the indicator which were assessed with the modelling
pathway. Again the cod indicator is most at risk, followed by the
modelled cod indicator, herring and seabirds. The pressures with the
highest negative impact are temperature and fishing. Zooplankton is an
example for an indicator which will likely thrive under the pressures.
It is still worthwhile to include such indicators in a risk assessment
as these positive changes in one indicator can have negative effects for
the entire system, thus every change can bear a risk for the system.
Overall the system is a bit more at risk considering the combination of
direct and indirect effects compared to direct effects only.

What can we do now with the results of our risk assessment? Here are
some examples:

-   We can inform the management by prioritizing which pressures pose
    the highest risk and which indicators are most at risk.
-   We can use the information to guide research: did any effect come by
    surprise (especially from the modelling pathway)? Which risks are
    associated with high uncertainties?
-   If we are interested in non-linear effects and how indicators and
    pressures interact with each other we could use our results as a
    basis for a network analysis and develop questions based on the risk
    assessment output or select specific indicator and pressure
    combinations that we want to further investigate with a more
    quantitative analysis.

# Contact

If you have issues with the package or questions please send an e-mail
to:
[helene.gutte\@uni-hamburg.de](mailto:helene.gutte@uni-hamburg.de){.email}

# References {#references}
